---
title: "MATH545_Assignment3_260677676_submitted"
author: "Dan Yunheum Seol"
date: '2018-11-20'
output:
  word_document: default
  pdf_document: default
---
#3.1
Among the time series process $\{X_t, t \in \mathbb{Z}\}$ of the form
\[
\phi_p(B)X_t = \theta_q(B)Z_t
\]
where
\[
\phi_p(z) =  1- \phi_1z - \phi_2z^2 - ... - \phi_pz^p
\]
and
\[
\theta_q(w) =  1+ \theta_1w - \theta_2w^2 + ... + \theta_qw^q
\]
Identify which are causal ot invertible
##(a)
\[
\phi_p(z) = 1+0.2z−0.48z^2 = 0
\]
is solved by (using quadratic formula) $z = \frac{-0.2\pm \sqrt{0.2^2 + 4*0.48}}{-0.48*2}$ 
\[z_1 = 5/3, z_2 = -5/4.\]
Thus it is causal.
\[
\theta_q(w) = 1
\]
which is a constant polynomial, so it does not have a root. So our condition for invertibility holds trivially. So it is invertible.

##(b)

\[
\phi_p(z) = 1+1.9z+0.88z^2 = 0
\]
is solved by (using quadratic formula) $z = \frac{-1.9\pm \sqrt{1.9^2 - 4*0.88}}{1.76}$ 
\[
z_1 = 10/11, z_2 = -5/4
\]
Because $|z_1| < 1$, it is not causal.

\[
\theta_q(w) = 1+0.2w+0.7w^2 =0
\]
has roots (by quadratic formula: I will omit the steps from now)
\[
w =−(1\pm i* 69)/7
\]
where $|w| = \sqrt{70/49} > 1$ for both case, so it is invertible.

##(c)
\[\phi(z) =  1+0.6z=0\]
has a solution $z=−5/3$, thus it is causal.
\[
\theta(w) = 1+1.2z = 0
\]
has a solution at $-5/6$, thus it is not invertible


## d) 
\[
\phi(z)=1+1.8z+0.81z^2 =0
\]
has unique solution of multiplicity 2, $z = -10/9$
Thus it is causal.
\[
\theta_q(w) = 1
\]
so it is invertible as well.
##e)
\[\phi(z)=1+1.6z=0\] has a solution $z= -5/8$
-Hence it is not causal.

\[
\theta_q(w) = 1−0.4w+0.04w^2 
\]
so it
has a solution $w=5$, which makes it invertible.



#3.9
##(a)
```{r}
#Importing possibly necessary packages
library(itsmr)
library(tidyverse)
library(forecast)
head(deaths)
class(deaths)
```
Now we will convert the data into a time series object:
```{r}
d_ts <- ts(deaths, frequency = 12, start = c(1973, 1))
#head(d_ts)
d_ts_nabla12 <- diff(d_ts, lag=12)
d_ts_nb1.12 <- d_ts_nabla12 %>% diff(., lag=1)

head(d_ts_nb1.12)
```
Now we will be finding sample mean, acvf, acf
```{r}

##sample mean
d_bar <- mean(d_ts_nb1.12)

##sample acvf
acvf <- acvf(d_ts_nb1.12, h = 20)
acvf

 
ggAcf(d_ts_nb1.12, lag=20)

acf_values

print("ACVF values are below:")

```
```{r}
innovations <- function(kappa, n){
  v <- rep(c(0), len =n+1)
  theta <- matrix(rep(c(0), len = n*n), nrow=n, ncol=n)
  
  v[1] = kappa[1]
  theta[1,1] = (1/v[1])*(kappa[1])
  v[2] = kappa[2] - theta[1,1]^2*v[1]
  theta[2,2] = (1/v[1])*(kappa[1])
 ##indices as entries : I know that theta_{nn} is not on Theta[n, n], but here I do so
  #for i will explicitly call them with entry indices
  #we will make a new matrix and allocate theta_{ij}'s at in the entry as 
  # it was presented in the book
  theta[2,1] = (1/v[2])*(kappa[2]-theta[1,1]*theta[2,2]*v[1])
  v[3] = kappa[1] - (theta[2,2]*v[1]+theta[2,1]*v[2])
  for(i in 3:n-1){
    theta[i,i] = (1/v[1])*(kappa[i+1])
    for(k in 1:i-1 ){
      s = 0
      for(j in 0: k-1){
      s = s + theta[k, k-j]*theta[i, i-j]*v[j+1]
      }
    
    theta[i, i-k] = (1/v[k+1])*(kappa[abs(i+1-k)] - s)
    }
   u =0
    for(j in 0:i-1){
      u =u + theta[i, i-j]*theta[i, i-j]*v[j+1]
    }
   v[i+1] =kappa(1) - u
  }
   theta_ <- matrix(rep(0, n*n), nrow=n, ncol=n)
   for(i in 2:n){
     for(j in 1:n-1){
       theta_[i,j] = theta[i-1, i-j] 
       ##theta_[2,1] = theta[1,1] i =2; j = 1
       ##theta_[3,1] = theta[2,2] i =3 j = 1, 2
     }
   }
   L <- list(THETA = theta_, V = v)
  return(L)
}
```

##(b)
```{r}
K <- rep(c(0), 13)
K[1] = acvf[1]
K[2] = acvf[2]
K[13] = acvf[13]
K

#I <- innovations(K, 12)
theta <- matrix(rep(c(0), len = n*n), nrow=n, ncol=n)
theta

```
So we can get the expression for the time series
\[
\{\nabla \nabla_{12}X_t\}
\]
which is
\[
\nabla \nabla_{12}X_t = Z_t + \theta_1Z_{t-1}
\]

where  $\theta_1 = -0.4963$

which is not invertible


#5.4
```{r}
R2 <- c(1, 0.427, 0.427, 1)
vec <- c(0.427, 0.475)
R2. <- matrix(R2, nrow=2, ncol=2)
#R2.
cor. <- matrix(vec, nrow=2, ncol=1)
R2.inv <- solve(R2.)
phi.hat <- R2.inv %*% cor.
phi.hat
############### calculating the sigma2hat#############
gamma0 <- 1.15

gamma<- c(0.427*1.15, 0.475*1.15)
gamma. <- matrix(gamma, nrow=2, ncol=1)
phiTgamma. <- t(phi.hat) %*% gamma.
phiTgamma.
phiTgamma.[1]
phiTgamma <- phiTgamma.[1]
sigma2hat <- gamma0-phiTgamma
sigma2hat
##############CI(mean)#####################
v. <- 1.15*(1+2*(0.427+0.475+0.169))

CI95 <- function(center, v, n){
  print(c(center-1.96*(sqrt(v)/sqrt(n)), center+1.96*(sqrt(v)/sqrt(n))))
}
CI95(3.82, v., 200 )
################CI(phi)######################
Gamma.hat <- c(1.15, 1.15*0.427, 1.15*0.427, 1.15)
Gamma.hat. <- matrix(Gamma.hat, nrow=2, ncol=2)
Gamma.hat
V.hat <- sigma2hat*solve(Gamma.hat.)
V.hat
v1 <- V.hat[1,1]
v2 <- V.hat[2,2]
CI95(phi.hat[1][1], v1, 200)
CI95(phi.hat[2,1], v2, 200)
c(0.274-0.139)
##################PACF######################

```
##(a), no

